---
title: "Define Crowding Treatment Groups"
author: "USN 3775"
format:
  html:
    self-contained: true
    toc: true
    toc-location: left
    toc-title: "Table of Contents"
    toc-numbering: true
    number-sections: true
editor_options: 
  chunk_output_type: console
---

```{=html}
<style>
body {
text-align: justify}
</style>
```

```{r, echo = FALSE}
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
options(scipen = 999)

library(librarian)
shelf(
  data.table,
  tidytable,
  quiet = TRUE
)
```

# Treatment Assignment

* In a hypothetical randomised controlled trial, patients are either allocated to a crowded or non-crowded ED.
* A crowded ED is defined as the Addenbrooke's ED at the top 20% of observed occupancy (80th or higher percentile of occupancy)
* A non-crowded ED is defined as the Addenbrooke's ED at the bottom 50% of observed occupancy (50th or lower percentile of occupancy)
* The remaining participants are used for sensitivity analyses.
* The crowding percentiles are re-calculated before assigning the treatment. They were calculates first in the pre-processing script. But there, the data included repeat ED visits. They are re-calculated here to reflect the distribution across patients' first visit.
* An alternative crowding definition is calculated as well, where only the top 10% are considered crowded.

```{r}
data <- readRDS("./data/processed/data_mids_firstvisit.RDS") |>
  mice::complete() |>
  as.data.table() |>
  select(-crowding_percentile) |>
  mutate(
    crowding_percentile = dplyr::ntile(ed_occupancy_at_arrival, 100),
    crowding_treatment = case_when(
      crowding_percentile >= 81 ~ "crowded",
      crowding_percentile <= 50 ~ "not_crowded",
      TRUE ~ "sensitivity"
    ),
    crowding_treatment_sens = case_when(
      crowding_percentile >= 91 ~ "crowded",
      crowding_percentile <= 60 ~ "not_crowded",
      TRUE ~ "sensitivity"
    ),
    crowding_treatment = as.factor(crowding_treatment),
    crowding_treatment_sens = as.factor(crowding_treatment_sens)
  )
```

# Transformation of Hours of Day

* First-order harmonic transformation of the hour-of-day to model it as cycle with length of 24 hours.
* A second order harmonic transformation resulted in higher AIC/BIC values when used in a logistic regression model with the binary outcome of crowding and was not used in subsequent analyses.

```{r}
data <- data |>
  mutate(
    arrival_hourmin = hour(ArrivalDateTime) + minute(ArrivalDateTime) / 60,
    hour_sin = sin(2 * pi * arrival_hourmin / 24),
    hour_cos = cos(2 * pi * arrival_hourmin / 24)
  )
```

# Data Export

```{r}
saveRDS(
  data,
  file = "./data/processed/data_tmle.RDS"
)
```
